### 2015-08-05 background.tex
Changed Figure into Fig.

Figure~\ref{fig:data_race_example} shows a racy entry point, \texttt{nvram\_llseek}, in the generic\_nvram Linux 4.0 driver. The driver can invoke the entry point concurrently from two threads, with the same \texttt{file} struct as a parameter. This can lead to multiple possible data races because the threads can access the \texttt{f\_pos} field of \texttt{file} in arbitrary order. Our tool, \whoop (see \S\ref{whoop}), was able to find these races automatically (see \S\ref{evaluation}).

@@

The most common method for avoiding races is by protecting sets of program statements that access a shared resource with \emph{locks}, forming \emph{critical sections}.  Figure~\ref{fig:lock_example} shows how to use locking to eliminate the races in Figure~\ref{fig:data_race_example}. Because the \texttt{return} statement can potentially race on the \texttt{f\_pos} field, we store the result in a temporary variable \texttt{res} inside the critical section.
%
Careless use of locks has many well-known pitfalls~\cite{sutter2005software}: coarse-grained locking can hurt performance as it limits the opportunity for concurrency, while fine-grained locking can easily lead to deadlocks. Although the Linux kernel provides sophisticated lock-free synchronization mechanisms~\cite[p.\ 123]{corbet2005linux}, such as atomic read-modify-write operations, in this work we focus on locks as they are widely used.\footnote{We treat lock-free operations soundly by regarding them as not providing any protection between threads; this can lead to false alarms.}

@@

Figure~\ref{fig:locksets} shows an example of applying lockset analysis to a concurrent program consisting of two threads $T_1$ and $T_2$, both accessing a global variable $A$. Initially, $\mathit{LS}_A$, which is the lockset for A, contains all possible locks in the program: $M$ and $N$. During execution of $T_1$, the thread writes $A$ without holding $N$, and thus $N$ is removed from $\mathit{LS}_A$. Next, during execution of $T_2$, the thread writes $A$ without holding $M$, and thus $\mathit{LS}_A$ becomes empty. As a result, a warning is reported because the two threads do not consistently protect $A$.

>>>>

Fig.~\ref{fig:data_race_example} shows a racy entry point, \texttt{nvram\_llseek}, in the generic\_nvram Linux 4.0 driver. The driver can invoke the entry point concurrently from two threads, with the same \texttt{file} struct as a parameter. This can lead to multiple possible data races because the threads can access the \texttt{f\_pos} field of \texttt{file} in arbitrary order. Our tool, \whoop (see \S\ref{whoop}), was able to find these races automatically (see \S\ref{evaluation}).

@@

The most common method for avoiding races is by protecting sets of program statements that access a shared resource with \emph{locks}, forming \emph{critical sections}.  Fig.~\ref{fig:lock_example} shows how to use locking to eliminate the races in Fig.~\ref{fig:data_race_example}. Because the \texttt{return} statement can potentially race on the \texttt{f\_pos} field, we store the result in a temporary variable \texttt{res} inside the critical section.
%
Careless use of locks has many well-known pitfalls~\cite{sutter2005software}: coarse-grained locking can hurt performance as it limits the opportunity for concurrency, while fine-grained locking can easily lead to deadlocks. Although the Linux kernel provides sophisticated lock-free synchronization mechanisms~\cite[p.\ 123]{corbet2005linux}, such as atomic read-modify-write operations, in this work we focus on locks as they are widely used.\footnote{We treat lock-free operations soundly by regarding them as not providing any protection between threads; this can lead to false alarms.}

@@

Fig.~\ref{fig:locksets} shows an example of applying lockset analysis to a concurrent program consisting of two threads $T_1$ and $T_2$, both accessing a global variable $A$. Initially, $\mathit{LS}_A$, which is the lockset for A, contains all possible locks in the program: $M$ and $N$. During execution of $T_1$, the thread writes $A$ without holding $N$, and thus $N$ is removed from $\mathit{LS}_A$. Next, during execution of $T_2$, the thread writes $A$ without holding $M$, and thus $\mathit{LS}_A$ becomes empty. As a result, a warning is reported because the two threads do not consistently protect $A$.

### 2015-08-05 evaluation.tex
Removed vspace hacks for now.
Changed Figure into Fig.

\newcommand{\colspacing}{\hspace{1.8em}}
\begin{table}[t]
\small
\centering
\setlength{\tabcolsep}{0.3em}
\caption{Program statistics and race-checking results from applying \whoop and \corral on our benchmarks.}
\label{tab:stats}
\begin{tabular}{l rrr rr r}
\centering
\input{experiments/tables/statistics.tex}
\end{tabular}
\vspace{-3mm}
\end{table}

\begin{table*}[t]
\small
\centering
\setlength{\tabcolsep}{0.4em}
\caption{Comparison with different yield instrumentation granularities and context-switch bounds.}
\label{tab:results}
\begin{tabular}{l r rrrr rrrr rrrr}
\centering
\input{experiments/tables/scalability.tex}
\end{tabular}
\vspace{-3mm}
\end{table*}

\begin{figure}
\centering
\includegraphics[width=.99\linewidth]{experiments/figures/yieldmr_vs_yieldall.pdf}
\vspace{-2mm}
\caption{Scatter plot showing the runtime speedups that \corral achieves using \whoop with the \yieldmr instrumentation. The symbols $+$, $\circ$ and $\times$, represent a context-switch bound of 2, 5 and 9, respectively.}
\label{fig:plot}
\vspace{-2mm}
\end{figure}

@@

As an example of the second category, we found the following race in the generic\_nvram driver (see Figure~\ref{fig:data_race_example}): the \texttt{llseek} entry point accesses the file offset \texttt{file->f\_pos} without a lock (\texttt{file} is passed as a parameter to \texttt{llseek}). This causes a race if the driver invokes \texttt{llseek} from another thread passing the same \texttt{file} object as a parameter.  We observed that another char driver, using the same APIs, \emph{does} use a mutex to protect the offset access in its \texttt{llseek} entry point, leading us to suspect that the race we found in generic\_nvram is not benign.  We have filed a bug report in the Linux kernel bug tracker and are awaiting a response.

\noindent\textbf{Accelerating \corral }
%
Table~\ref{tab:results} presents runtime results from using \whoop, \corral and \whoop + \corral to analyze our benchmarks, while Figure~\ref{fig:plot} plots the runtime speedups that \corral achieves using \whoop with the \yieldmr instrumentation. All reported times are in seconds and averaged over three runs. \corral was configured with a time budget of 10 hours (T.O. denotes timeout) and a csb of 2, 5 and 9. Standalone \corral uses \yieldall, which instruments context-switches (i.e. \texttt{yield} statements) after all visible operations, while \whoop + \corral uses \yieldcoarse and \yieldmr, which instrument context-switches in a more fine-grained fashion (see \S\ref{corral}). The table also shows the number of context-switches per instrumentation (\#Y).

@@

Using the race-freedom guarantees from \whoop, we managed to significantly accelerate \corral in most of our benchmarks; the best results were achieved using \yieldmr. Figure~\ref{fig:plot} shows that most speedups using \yieldmr are between 1.5$\times$ and 10$\times$; while in ssu100, and pc8736x\_gpio respectively, we achieved a speedup of 12$\times$, and 20$\times$ respectively, with a csb of 9. We noticed that a higher csb typically results into greater speedups when exploiting \whoop. This is expected, as a higher csb translates into a larger sequentialized program, and thus \whoop has more opportunities to reduce this sequentialization.
%
However, there are some cases where \whoop might slow down \corral. We noticed this in the sx8 driver: \whoop verified 46 out of 47 memory regions, but did not fully verify any of the pairs; \corral, on the other hand, analyzed the only two pairs of the driver in just 21.4 seconds (csb of 9). We believe that in this case the overhead of running \whoop outweighed the benefits of using \yieldmr.

>>>>

\newcommand{\colspacing}{\hspace{1.8em}}
\begin{table}[t]
\small
\centering
\setlength{\tabcolsep}{0.3em}
\caption{Program statistics and race-checking results from applying \whoop and \corral on our benchmarks.}
\label{tab:stats}
\begin{tabular}{l rrr rr r}
\centering
\input{experiments/tables/statistics.tex}
\end{tabular}
%\vspace{-3mm}
\end{table}

\begin{table*}[t]
\small
\centering
\setlength{\tabcolsep}{0.4em}
\caption{Comparison with different yield instrumentation granularities and context-switch bounds.}
\label{tab:results}
\begin{tabular}{l r rrrr rrrr rrrr}
\centering
\input{experiments/tables/scalability.tex}
\end{tabular}
%\vspace{-3mm}
\end{table*}

\begin{figure}
\centering
\includegraphics[width=.99\linewidth]{experiments/figures/yieldmr_vs_yieldall.pdf}
%\vspace{-2mm}
\caption{Scatter plot showing the runtime speedups that \corral achieves using \whoop with the \yieldmr instrumentation. The symbols $+$, $\circ$ and $\times$, represent a context-switch bound of 2, 5 and 9, respectively.}
\label{fig:plot}
%\vspace{-2mm}
\end{figure}

@@

As an example of the second category, we found the following race in the generic\_nvram driver (see Fig.~\ref{fig:data_race_example}): the \texttt{llseek} entry point accesses the file offset \texttt{file->f\_pos} without a lock (\texttt{file} is passed as a parameter to \texttt{llseek}). This causes a race if the driver invokes \texttt{llseek} from another thread passing the same \texttt{file} object as a parameter.  We observed that another char driver, using the same APIs, \emph{does} use a mutex to protect the offset access in its \texttt{llseek} entry point, leading us to suspect that the race we found in generic\_nvram is not benign.  We have filed a bug report in the Linux kernel bug tracker and are awaiting a response.

\noindent\textbf{Accelerating \corral }
%
Table~\ref{tab:results} presents runtime results from using \whoop, \corral and \whoop + \corral to analyze our benchmarks, while Fig.~\ref{fig:plot} plots the runtime speedups that \corral achieves using \whoop with the \yieldmr instrumentation. All reported times are in seconds and averaged over three runs. \corral was configured with a time budget of 10 hours (T.O. denotes timeout) and a csb of 2, 5 and 9. Standalone \corral uses \yieldall, which instruments context-switches (i.e. \texttt{yield} statements) after all visible operations, while \whoop + \corral uses \yieldcoarse and \yieldmr, which instrument context-switches in a more fine-grained fashion (see \S\ref{corral}). The table also shows the number of context-switches per instrumentation (\#Y).

@@

Using the race-freedom guarantees from \whoop, we managed to significantly accelerate \corral in most of our benchmarks; the best results were achieved using \yieldmr. Fig.~\ref{fig:plot} shows that most speedups using \yieldmr are between 1.5$\times$ and 10$\times$; while in ssu100, and pc8736x\_gpio respectively, we achieved a speedup of 12$\times$, and 20$\times$ respectively, with a csb of 9. We noticed that a higher csb typically results into greater speedups when exploiting \whoop. This is expected, as a higher csb translates into a larger sequentialized program, and thus \whoop has more opportunities to reduce this sequentialization.
%
However, there are some cases where \whoop might slow down \corral. We noticed this in the sx8 driver: \whoop verified 46 out of 47 memory regions, but did not fully verify any of the pairs; \corral, on the other hand, analyzed the only two pairs of the driver in just 21.4 seconds (csb of 9). We believe that in this case the overhead of running \whoop outweighed the benefits of using \yieldmr.

### 2015-08-05 whoop.tex
Changed Figure into Fig.

The allowed statements are shown in Figure~\ref{fig:statements}, and include designated statements for reading from and writing to shared variables.  In particular, shared variables may not appear in arbitrary expressions.  This restriction simplifies our presentation of lockset instrumentation below, and a program that does not satisfy this restriction can be trivially pre-processed into one that does via the introduction of additional private temporary variables to record values read from the shared state.  We do not specify the form of expressions, nor the types of variables, assuming a standard set of data types and operations.

@@

Lockset instrumentation introduces: sets $\hasbeenread{i} \subseteq \powerset{\sharedvars}$ and $\hasbeenwritten{i} \subseteq \powerset{\sharedvars}$ to track the shared variables that have been read from and written to, respectively, by the thread executing $T$; a current lockset $\currentlockset{i} \subseteq \powerset{\mutexes}$ to record the mutexes currently held by the thread; and, for each shared variable $s \in \sharedvars$, a lockset $\lockset{s}{i}$ to record the mutexes that are consistently held when the thread accesses $s$.
%
The statements of each procedure in $\instrument{T}{i}$ that access shared variables and mutexes are instrumented to manipulate these sets, as shown in Figure~\ref{fig:instrumentation}.  For a shared variable assignment $s = e$, we record in $\hasbeenwritten{i}$ that $s$ has been written to, and update $\lockset{s}{i}$ to eliminate any mutexes that are not currently held (those mutexes that are not in $\currentlockset{i}$).  A shared variable read $x = s$ is instrumented analogously, with an additional $\havockeyword$ command which we discuss below.  Instrumentation of mutex manipulation commands, $\mutexlock{m}$ and $\mutexunlock{m}$, involves updating $\currentlockset{i}$ to add and remove mutex $m$, respectively.

\noindent\textbf{Shared State Abstraction }
%
Recall that while our aim is to perform race analysis for pairs of threads, we must be sure to account for possible side-effects due to other threads that are running concurrently.  The instrumentation of Figure~\ref{fig:instrumentation} achieves this via \emph{nondeterminism}: when reading from a shared variable $s$, a nondeterministic value is returned.  This is reflected by the use of a $\havockeyword$ command, which sets its argument to an arbitrary value.  Because all shared state accesses are abstracted in this fashion, it is possible to completely dispense with the shared variables after the lockset instrumentation has been performed.  As a result, when instrumenting a shared variable write, the effect of the write is not explicitly modeled.

\noindent\textbf{Sequentialization }
%
The pseudocode of Figure~\ref{fig:sequentialization} shows the sequential program that we analyze in order to prove race-freedom for a pair of thread templates $T$ and $U$. Assuming that $T$ and $U$ have been instrumented using distinct symbols $i$ and $j$, yielding $\instrument{T}{i}$ and $\instrument{U}{j}$, the sequential program operates as follows. First, the read, write, and current locksets for $\instrument{T}{i}$ and $\instrument{U}{j}$ are initialized to be empty, and for each shared variable $s$, the locksets $\lockset{s}{i}$ and $\lockset{s}{j}$ are initialized to the full set of mutexes $\mutexes$.  The main procedures of the instrumented thread templates, $\mainforthread{\instrument{T}{i}}$ and $\mainforthread{\instrument{U}{j}}$, are then executed in turn (the order does not matter due to renaming). Finally, an assertion checks for consistent use of mutexes: if $s$ is written during execution of $\instrument{T}{i}$ and accessed during execution of $\instrument{U}{j}$, or vice-versa, then the locksets $\lockset{s}{i}$ and $\lockset{s}{j}$ must contain at least one common mutex.

@@

\noindent\textbf{Soundness }
%
We sketch an argument that if the program of Figure~\ref{fig:sequentialization} is correct (i.e.\ the assertion, described above, holds at the end of the program), then it is impossible for a thread executing template $T$ to race with a thread executing template $U$, under the assumption that the threads are guaranteed to terminate. Let us assume that the program of Figure~\ref{fig:sequentialization} is correct, and suppose (by way of contradiction) that a thread executing $T$ can in fact race with a thread executing $U$, on some shared variable $s$.  By our hypothesis that the program is correct, and that the threads terminate, the assertion checked at the end of the program guarantees that at least one mutex, say $m$, belongs to both $\lockset{s}{i}$ and $\lockset{s}{j}$.  By the definition of a lockset (and according to the manner in which shared accesses are instrumented in Figure~\ref{fig:instrumentation}), this means that $m$ is held during every access to $s$ by both $\instrument{T}{i}$ and $\instrument{U}{j}$. As a result, $m$ must be unlocked and locked between the two accesses, which contradicts that the pair of accesses is racing.

In the presence of non-termination the assertion at the end of Figure~\ref{fig:sequentialization} may not be reached.  The termination analysis problem for device drivers has been widely studied (see e.g.\ ~\cite{cook2006termination}), and in the remainder of the paper we do not consider termination issues, assuming that the drivers we analyze in our experimental evaluation (see \S\ref{evaluation}) are terminating.

@@

The simple concurrent programming model of \S\ref{sec:symbolicpairwise} is deliberately idealistic to make it easy to describe our symbolic verification technique. In practice, Linux drivers are written in C, our technique does not know up-front which are the driver entry points, drivers do not work with a cleanly specified set of named locks, and rather than having a given set of named shared variables, we have arbitrary memory accesses via pointers. We now explain how we have taken the conceptual ideas from \S\ref{sec:symbolicpairwise} and used them to build \whoop,\footnote{\url{https://github.com/pdeligia/whoop}} a practical, fully automatic tool for detecting data races in drivers.

\noindent\textbf{Architecture }
%
Figure~\ref{fig:whoop} depicts the \whoop toolchain. The input to \whoop is a Linux driver written in C, together with an environmental model\footnote{This consists of stub header files modeling relevant Linux kernel APIs.} that is required to ``close'' the driver so that it can be analyzed for races. Initially, \whoop uses three LLVM\footnote{\url{http://llvm.org}}-based tools, Chauffeur,\footnote{\url{https://github.com/mc-imperial/chauffeur}} Clang\footnote{\url{http://clang.llvm.org}} and SMACK\footnote{\url{https://github.com/smackers/smack}}~\cite{rakamaric2014smack}, to translate the driver and its environmental model into an abstract program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}, a simple imperative language with well-defined semantics that is used as the input to a number of cutting-edge verifiers (e.g.\ Boogie and \corral). Next, \whoop instruments and sequentializes the program to perform symbolic pairwise lockset analysis (see Figure~\ref{fig:whoop} -- B and \S\ref{sec:symbolicpairwise}) using the Boogie verification engine. After the verification phase ends, \whoop can exploit any inferred race-freedom guarantees to accelerate precise race-checking with \corral (see Figure~\ref{fig:whoop} -- C and \S\ref{corral}).

We engineered the Chauffeur and \whoop components of our toolchain (denoted with grey boxes in Figure~\ref{fig:whoop}).  For the remaining components, we were able to reuse industrial-strength tools that are robust and battle-proven via their use in many complex software projects.

\noindent\textbf{Extracting Entry Point Information }
%
Chauffeur is a Clang-based tool that traverses the abstract syntax tree (AST) of the original driver and extracts all entry point identifier names, together with the identifier names of their corresponding kernel functions. Linux drivers define entry points in a standard way (see Figure~\ref{fig:entrypoints} for an example of how the generic\_nvram driver defines the entry points for \texttt{file\_operations}); Chauffeur identifies these definitions in the AST and outputs the relevant information in an XML file, which is parsed by \whoop to be used during the instrumentation.

@@

\noindent\textbf{Watchdog Race-Checking Instrumentation }
%
Data race detection is performed by introducing sets containing the locks that are consistently held when accessing each shared variable and sets containing all shared variables that are read and written (see \S\ref{sec:symbolicpairwise} and the instrumentation of Figure~\ref{fig:instrumentation}). These sets can be modeled directly in Boogie as characteristic functions, using maps. However, this requires the use of quantifiers to express properties related to set contents.  For instance, to express that a set $X$ of elements of type $A$ is empty, where $X$ is represented as a map from $A$ to $\mathsf{Bool}$, we would require the quantified expression $\forall a : A : \neg X[a]$.  It is well known that automated theorem proving in the presence of quantified constraints is challenging, and that theorem provers such as Z3~\cite{de2008z3} are often much more effective when quantifiers are avoided.

@@

\noindent\textbf{Verification and Error Reporting }
%
For each entry point pair, the instrumented sequential program, equipped with procedure and loop summaries, is sent to the Boogie verification engine. For each procedure in the program, Boogie generates a VC and discharges it to the Z3 theorem prover.  In particular, the verification for the root-level procedure, encoding the sequential program sketched in Figure~\ref{fig:sequentialization}, encodes the race-freedom check for the entry point pair. Successful verification implies that the pair is race-free, while an error (i.e.\ counterexample) denotes a \emph{potential} data race and is reported to the user. To improve usability, \whoop has a built-in error reporter that matches counterexamples to source code. The following is a race that \whoop found and reported for the example of Figure~\ref{fig:data_race_example}:

>>>>

The allowed statements are shown in Fig.~\ref{fig:statements}, and include designated statements for reading from and writing to shared variables.  In particular, shared variables may not appear in arbitrary expressions.  This restriction simplifies our presentation of lockset instrumentation below, and a program that does not satisfy this restriction can be trivially pre-processed into one that does via the introduction of additional private temporary variables to record values read from the shared state.  We do not specify the form of expressions, nor the types of variables, assuming a standard set of data types and operations.

@@

Lockset instrumentation introduces: sets $\hasbeenread{i} \subseteq \powerset{\sharedvars}$ and $\hasbeenwritten{i} \subseteq \powerset{\sharedvars}$ to track the shared variables that have been read from and written to, respectively, by the thread executing $T$; a current lockset $\currentlockset{i} \subseteq \powerset{\mutexes}$ to record the mutexes currently held by the thread; and, for each shared variable $s \in \sharedvars$, a lockset $\lockset{s}{i}$ to record the mutexes that are consistently held when the thread accesses $s$.
%
The statements of each procedure in $\instrument{T}{i}$ that access shared variables and mutexes are instrumented to manipulate these sets, as shown in Fig.~\ref{fig:instrumentation}.  For a shared variable assignment $s = e$, we record in $\hasbeenwritten{i}$ that $s$ has been written to, and update $\lockset{s}{i}$ to eliminate any mutexes that are not currently held (those mutexes that are not in $\currentlockset{i}$).  A shared variable read $x = s$ is instrumented analogously, with an additional $\havockeyword$ command which we discuss below.  Instrumentation of mutex manipulation commands, $\mutexlock{m}$ and $\mutexunlock{m}$, involves updating $\currentlockset{i}$ to add and remove mutex $m$, respectively.

\noindent\textbf{Shared State Abstraction }
%
Recall that while our aim is to perform race analysis for pairs of threads, we must be sure to account for possible side-effects due to other threads that are running concurrently.  The instrumentation of Fig.~\ref{fig:instrumentation} achieves this via \emph{nondeterminism}: when reading from a shared variable $s$, a nondeterministic value is returned.  This is reflected by the use of a $\havockeyword$ command, which sets its argument to an arbitrary value.  Because all shared state accesses are abstracted in this fashion, it is possible to completely dispense with the shared variables after the lockset instrumentation has been performed.  As a result, when instrumenting a shared variable write, the effect of the write is not explicitly modeled.

\noindent\textbf{Sequentialization }
%
The pseudocode of Fig.~\ref{fig:sequentialization} shows the sequential program that we analyze in order to prove race-freedom for a pair of thread templates $T$ and $U$. Assuming that $T$ and $U$ have been instrumented using distinct symbols $i$ and $j$, yielding $\instrument{T}{i}$ and $\instrument{U}{j}$, the sequential program operates as follows. First, the read, write, and current locksets for $\instrument{T}{i}$ and $\instrument{U}{j}$ are initialized to be empty, and for each shared variable $s$, the locksets $\lockset{s}{i}$ and $\lockset{s}{j}$ are initialized to the full set of mutexes $\mutexes$.  The main procedures of the instrumented thread templates, $\mainforthread{\instrument{T}{i}}$ and $\mainforthread{\instrument{U}{j}}$, are then executed in turn (the order does not matter due to renaming). Finally, an assertion checks for consistent use of mutexes: if $s$ is written during execution of $\instrument{T}{i}$ and accessed during execution of $\instrument{U}{j}$, or vice-versa, then the locksets $\lockset{s}{i}$ and $\lockset{s}{j}$ must contain at least one common mutex.

@@

\noindent\textbf{Soundness }
%
We sketch an argument that if the program of Fig.~\ref{fig:sequentialization} is correct (i.e.\ the assertion, described above, holds at the end of the program), then it is impossible for a thread executing template $T$ to race with a thread executing template $U$, under the assumption that the threads are guaranteed to terminate. Let us assume that the program of Fig.~\ref{fig:sequentialization} is correct, and suppose (by way of contradiction) that a thread executing $T$ can in fact race with a thread executing $U$, on some shared variable $s$.  By our hypothesis that the program is correct, and that the threads terminate, the assertion checked at the end of the program guarantees that at least one mutex, say $m$, belongs to both $\lockset{s}{i}$ and $\lockset{s}{j}$.  By the definition of a lockset (and according to the manner in which shared accesses are instrumented in Fig.~\ref{fig:instrumentation}), this means that $m$ is held during every access to $s$ by both $\instrument{T}{i}$ and $\instrument{U}{j}$. As a result, $m$ must be unlocked and locked between the two accesses, which contradicts that the pair of accesses is racing.

In the presence of non-termination the assertion at the end of Fig.~\ref{fig:sequentialization} may not be reached.  The termination analysis problem for device drivers has been widely studied (see e.g.\ ~\cite{cook2006termination}), and in the remainder of the paper we do not consider termination issues, assuming that the drivers we analyze in our experimental evaluation (see \S\ref{evaluation}) are terminating.

@@

The simple concurrent programming model of \S\ref{sec:symbolicpairwise} is deliberately idealistic to make it easy to describe our symbolic verification technique. In practice, Linux drivers are written in C, our technique does not know up-front which are the driver entry points, drivers do not work with a cleanly specified set of named locks, and rather than having a given set of named shared variables, we have arbitrary memory accesses via pointers. We now explain how we have taken the conceptual ideas from \S\ref{sec:symbolicpairwise} and used them to build \whoop,\footnote{\url{https://github.com/pdeligia/whoop}} a practical, automated tool for detecting data races in drivers.

\noindent\textbf{Architecture }
%
Fig.~\ref{fig:whoop} depicts the \whoop toolchain. The input to \whoop is a Linux driver written in C, together with an environmental model\footnote{This consists of stub header files modeling relevant Linux kernel APIs.} that is required to ``close'' the driver so that it can be analyzed for races. Initially, \whoop uses three LLVM\footnote{\url{http://llvm.org}}-based tools, Chauffeur,\footnote{\url{https://github.com/mc-imperial/chauffeur}} Clang\footnote{\url{http://clang.llvm.org}} and SMACK\footnote{\url{https://github.com/smackers/smack}}~\cite{rakamaric2014smack}, to translate the driver and its environmental model into an abstract program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}, a simple imperative language with well-defined semantics that is used as the input to a number of cutting-edge verifiers (e.g.\ Boogie and \corral). Next, \whoop instruments and sequentializes the program to perform symbolic pairwise lockset analysis (see Fig.~\ref{fig:whoop}--B and \S\ref{sec:symbolicpairwise}) using the Boogie verification engine. After the verification phase ends, \whoop can exploit any inferred race-freedom guarantees to accelerate precise race-checking with \corral (see Fig.~\ref{fig:whoop}--C and \S\ref{corral}).

We engineered the Chauffeur and \whoop components of our toolchain (denoted with grey boxes in Fig.~\ref{fig:whoop}).  For the remaining components, we were able to reuse industrial-strength tools that are robust and battle-proven via their use in many complex software projects.

\noindent\textbf{Extracting Entry Point Information }
%
Chauffeur is a Clang-based tool that traverses the abstract syntax tree (AST) of the original driver and extracts all entry point identifier names, together with the identifier names of their corresponding kernel functions. Linux drivers define entry points in a standard way (see Fig.~\ref{fig:entrypoints} for an example of how the generic\_nvram driver defines the entry points for \texttt{file\_operations}); Chauffeur identifies these definitions in the AST and outputs the relevant information in an XML file, which is parsed by \whoop to be used during the instrumentation.

@@

\noindent\textbf{Watchdog Race-Checking Instrumentation }
%
Data race detection is performed by introducing sets containing the locks that are consistently held when accessing each shared variable and sets containing all shared variables that are read and written (see \S\ref{sec:symbolicpairwise} and the instrumentation of Fig.~\ref{fig:instrumentation}). These sets can be modeled directly in Boogie as characteristic functions, using maps. However, this requires the use of quantifiers to express properties related to set contents.  For instance, to express that a set $X$ of elements of type $A$ is empty, where $X$ is represented as a map from $A$ to $\mathsf{Bool}$, we would require the quantified expression $\forall a : A : \neg X[a]$.  It is well known that automated theorem proving in the presence of quantified constraints is challenging, and that theorem provers such as Z3~\cite{de2008z3} are often much more effective when quantifiers are avoided.

@@

\noindent\textbf{Verification and Error Reporting }
%
For each entry point pair, the instrumented sequential program, equipped with procedure and loop summaries, is sent to the Boogie verification engine. For each procedure in the program, Boogie generates a VC and discharges it to the Z3 theorem prover.  In particular, the verification for the root-level procedure, encoding the sequential program sketched in Fig.~\ref{fig:sequentialization}, encodes the race-freedom check for the entry point pair. Successful verification implies that the pair is race-free, while an error (i.e.\ counterexample) denotes a \emph{potential} data race and is reported to the user. To improve usability, \whoop has a built-in error reporter that matches counterexamples to source code. The following is a race that \whoop found and reported for the example of Fig.~\ref{fig:data_race_example}:
